Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-19 rebuttal.tex

(6) There is only limited statistical assessment/validation of the models. Statements made in the
text that R2 values are “reasonable” and the models perform “relatively well” are difficult to
reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.

(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are
not relevant for activity cliff assessment), SALI values are consistently over-predicted. By
contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in
part significantly. This is not investigated, although the apparent errors at high value ranges are
the perhaps most critical features of the reported RF models.

(8) In the section “Extending a landscape”, predictions are finally reported for only three (!)
molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis
of the results reported for this little bit of evaluation, one would be hard pressed to make a case
for the ability of the approach to predict activity cliffs with any certainty.

\textbf{(9) For the prediction of activity cliffs, SALI values must be calculated for compound
  pairs. This is the central idea of the approach. Accordingly, fingerprint descriptors and potency
  values must be combined in some ways.  \vskip 1em (9.1.) It is nowhere stated how activity values
  are treated for compounds forming a pair. Are activity differences calculated at the pair level?
  If so, how are activity differences compared for the calculation of pair-based SALI scores - as a
  difference of differences? It should be noted predicted potency values in Figures 4-6 are
  essentially all over the place.}

The entire premise of the proposed method is that one evaluates SALI values for a dataset that has
observed activities. We state specifically (Equation 1) how the activities are combined when
molecules are considered pairwise.

@@

It's not clear what the reviewer means by \emph{``\ldots how are activity differences compared for
  the calculation of pair-based SALI scores \ldots''}. The SALI value itself represents a pair of
molecules and in this calculation, we consider the absolute activity difference. We do not consider
pairs of pairs

\textbf{(9.2.) To obtain a fingerprint descriptor for a pair, the author aggregates fingerprints of
  individual molecules, finally, by simple averaging}

In fact, we do not generate pairwise fingerprint descriptors. Fingerprints are only used to evaluate
the similarity term in the calculation of the SALI value.

\textbf{For comparisons of pairs, Tanimoto similarity is then calculated for averaged fingerprints,
  which presents an artificial assessment of pair similarity (one can easily come up with a few
  hypothetical compound similarity relationships that make this assessment questionable at best). As
  a consequence, activity cliffs are considered for which Tanimoto similarity calculations yield
  values between 0.2 and 0.3!}

@@

I believe that these comments are based on the discussion in Section 3.1 and it appears that this
section of the paper was not clear.

Firstly, we have never mentioned the averaging of fingerprint descriptors. Rather, the averaging is
applied to the topological descritpors (i.e., the X variables). Specifically, we generate a
descriptor vector for a pair of molecules by averaging the (topological) descriptor vectors of the
individual members of that pair.

@@

The goal of Section 3.1 is to indicate that even though the Y variable and the X variables encode
structural information (the former indirectly and the latter directly), there is little correlation
between the individual (aggregated, topological) descriptors and the Y variable (where sructural
information is included via fingerprints). To quantify this, we evaluate the Pearson correlation
between each (aggregated, topological) descriptor and the Y variable and show that the maximum value
is very low ($R^2 < 0.15$).

This discussion does not involve fingerprint-based Tanimoto similarities at all. In fact, if one
does indeed consider the pairwise Tanimoto values of the molecules within a dataset, we do in fact
observe an appreciable number of pairs wth $T_c > 0.8$.

\textbf{Thus, on the basis of these similarity calculations, one could never decide which level of
  similarity might be cliff relevant. In this respect, the descriptions of activity cliff by the
  author are quite telling – he speaks of “predicted” versus “true” activity cliffs, of “relatively
  accurate” activity cliffs, and of “most significant activity cliffs that are in fact not very
  significant activity cliffs in an absolute sense”, and so on. }

@@

I agree that the wording of this discussion is rather sloppy. The text has been updated to be more
quantitative.

Yet at the same time, the concept of an activity cliff is somewhat subjective. In many cases, it is
akin to saying \emph{``I'll know it's an activity cliff when I see it''}. For example, a pair of
molecules exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously an activity
clif. Would a pair with $T_C = 0.85$ and 50x difference in activity not be an activity cliff? From
this point of view, I do not think it unreasonable to refer somewhat qualitatively to the
``cliffness'' of an activity cliff.

\textbf{Furthermore, there might be many instances of activity and fingerprint similarity
  relationships between pairs of compounds that might yield high pair-based SALI scores dominated by
  either the activity or similarity component, without forming a “true” activity cliff. The
  conclusion of this reviewer is that it is currently not possible to capture activity cliffs with
  any degree of certainty on the basis of the pair calculations, as reported.}

\textbf{(10) It also remains largely obscure how an activity landscape should be extended following
  this approach by compounds for which no activity information is available. Should the activity
  first be predicted by an RF model and then used for pair-based SALI score calculations?  One would
  hope not.}

The initial description was actually misleading. The prediction of SALI values permits us to first
extend the SALI matrix. As correctly noted by the reviewer, this does not lead to an extension of
the structure-activity landscape. In the proposed methodology, the SALI value for a new molecule is
always predicted \emph{with respect to} another molecule whose activity is known. Thus, given a
predicted SALI value, one can rearrange Equation 1 to obtain a ``predicted'' activity value for the
test molecule. This value can be used to extend the original landscape. The text has been updated to
discuss this correctly XXX

Of course, if one predicts a SALI value for a pair of molecules, in which both members have an
unknown actvity, one cannot extend the landscape.

### 2012-07-20 rebuttal.tex

Yet at the same time, the concept of an activity cliff is somewhat subjective. In many cases, it is
akin to saying \emph{``I'll know it's an activity cliff when I see it''}. For example, a pair of
molecules exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously an activity
clif. Would a pair with $T_C = 0.85$ and 50x difference in activity not be an activity cliff? From
this point of view, I do not think it unreasonable to refer somewhat qualitatively to the
``cliffness'' of an activity cliff.

\textbf{Furthermore, there might be many instances of activity and fingerprint similarity
  relationships between pairs of compounds that might yield high pair-based SALI scores dominated by
  either the activity or similarity component, without forming a “true” activity cliff. The
  conclusion of this reviewer is that it is currently not possible to capture activity cliffs with
  any degree of certainty on the basis of the pair calculations, as reported.}

\textbf{(10) It also remains largely obscure how an activity landscape should be extended following
  this approach by compounds for which no activity information is available. Should the activity
  first be predicted by an RF model and then used for pair-based SALI score calculations?  One would
  hope not.}

The initial description was in misleading. In fact, the approach does not allow one to obtain
predicted activity values, due to the fact that the sign of the difference in activities in the SALI
formula is ignored. While one could create another model in parallel, that predicts signed activity
differences (and thereby determine whether the activity derived from the predicted SALI value should
be less than or greater than that of the molecule with observed activity), this would probably not
be very reliable. The manuscript has been updated to remove the dicussion on extending SAR
landscapes.

### 2012-07-20 rebuttal.tex

Firstly, we have never mentioned the averaging of fingerprint descriptors. Rather, the averaging is
applied to the topological descritpors (i.e., the X variables). Specifically, we generate a
descriptor vector for a pair of molecules by averaging the (topological) descriptor vectors of the
individual members of that pair.

The goal of Section 3.1 is to indicate that even though the Y variable and the X variables encode
structural information (the former indirectly and the latter directly), there is little correlation
between the individual (aggregated, topological) descriptors and the Y variable (where sructural
information is included via fingerprints). To quantify this, we evaluate the Pearson correlation
between each (aggregated, topological) descriptor and the Y variable and show that the maximum value
is very low ($R^2 < 0.15$).

This discussion does not involve fingerprint-based Tanimoto similarities at all. In fact, if one
does indeed consider the pairwise Tanimoto values of the molecules within a dataset, we do in fact
observe an appreciable number of pairs wth $T_c > 0.8$.

In refering to pairs with $T_C$ values between 0.2 and 0.3 -- it is true that such pairs are not
really activity cliffs in the original sense of the term. As described below, defining a threshold
SALI, above which a pair represents an activity cliff, is rather subjective.

@@

\textbf{Thus, on the basis of these similarity calculations, one could never decide which level of
  similarity might be cliff relevant. In this respect, the descriptions of activity cliff by the
  author are quite telling – he speaks of “predicted” versus “true” activity cliffs, of “relatively
  accurate” activity cliffs, and of “most significant activity cliffs that are in fact not very
  significant activity cliffs in an absolute sense”, and so on. }

I agree that the wording of this discussion is somewhat sloppy. The text has been updated to be more
quantitative.

Yet at the same time, the concept of an activity cliff is subjective. In many cases, it is
akin to saying \emph{``I'll know it's an activity cliff when I see it''}. For example, a pair of
molecules exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously an activity
clif. Would a pair with $T_C = 0.85$ and 50x difference in activity not be an activity cliff? From
this point of view, I do not think it unreasonable to refer somewhat qualitatively to the
``cliffness'' of an activity cliff.

Certainly, with respect to the previous comment, a pair of molecules with $T_c = 0.29$ would likely
not be regarded as a cliff - I specifically note that such a case is not a significant cliff. To
address this aspect of SALI values, I have presented plots of dissimilarity versus activity ratio
(Figures 7 and 9) which allow one to more directly determine ``significant'' versus
``insignificant'' cliffs.

@@

\textbf{Furthermore, there might be many instances of activity and fingerprint similarity
  relationships between pairs of compounds that might yield high pair-based SALI scores dominated by
  either the activity or similarity component, without forming a “true” activity cliff. The
  conclusion of this reviewer is that it is currently not possible to capture activity cliffs with
  any degree of certainty on the basis of the pair calculations, as reported.}

\textbf{(10) It also remains largely obscure how an activity landscape should be extended following
  this approach by compounds for which no activity information is available. Should the activity
  first be predicted by an RF model and then used for pair-based SALI score calculations?  One would
  hope not.}

The original description was misleading. In fact, the approach does not allow one to obtain
predicted activity values, due to the fact that the sign of the difference in activities in the SALI
formula is ignored. While one could create another model in parallel, that predicts signed activity
differences (and thereby determine whether the activity derived from the predicted SALI value should
be less than or greater than that of the molecule with observed activity), this would probably not
be very reliable! The manuscript has been updated to remove the dicussion on extending SAR
landscapes.

### 2012-07-20 paper.tex
Added more rebuttal
removed discussion of extending landscapes. Updated rebuttal
more fixes

The preceding discussion raises the issue of explicitly duplicating structural information in the
independent and dependent variables. More specifically, both the dependent variable (i.e., the SALI
values) and the aggregated descriptors characterize the molecules' structure. One might therefore
ask whether models built using such data perform over-optimistically. Given the nature of the
descriptors used in the independent and dependent variables (the former being based on multiple atom
and bond features and the latter derived from purely topological paths), we believe that such a
phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram of the pairwise Pearson
correlations between the SALI values and each of the descriptor values for the
Cavalli\cite{Cavalli:2002aa} dataset, using the different aggregation functions described above. It
is evident, that the highest $R^2$, between any of the descriptors and the SALI values is less than
0.15. Similar behavior was observed for all the datasets used in this study. These observations
suggest that problems arising from the inclusion of structural information simultaneously in the
dependent and independent variables is minimal.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
\ref{tab:rfmodel-results} reports the model statistics.  While the differences in model performance
do not vary significantly with the aggregation function used, we see that for the Costanzo and Kalla
datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and $R^2$ than when
using $f_{\textrm{mean}}$.

Given the relatively small differences between models based on different aggregation functions, we
arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function for subsequent
analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred} summarize the
performance of the models built using the reported activity data and the pairwise SALI values for
the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, the most significant cliffs
exhibit significant variance. In comparison, the models developed on the reported activity values
are relatively poor. This can be partially ascribed to the small sizes of the dataset. It should be
note that this methodology is not aimed at modeling the original, usually small, SAR dataset. It is
well known (and evident from the current analyses) that small datasets do not lead to very robust or
reliable machine learning models. Instead, the methodology focuses on the \emph{pairwise} dataset --
which are usually large enough for a statistically significant model.

\subsection{Extending a landscape}
\label{sec:extending-landscape}

While predictive modeling of the SALI values is a first step, the primary goal of this approach is
to extend the structure-activity landscape for a set of molecules. In other words, can one use the
model to identify compounds that will exhibit an activity cliff with one or more molecules in the
dataset? Given a new molecule, one can evaluate the descriptors for this molecule and then evaluate
the pairwise descriptor values with each member of the training set. Thus, given a new molecule, we
must obtain predictions for $n$ ``pairwise objects'', where $n$ is the number of molecules in the
training set. Given these $n$ predictions one can proceed to identify training set molecules with
which the new molecule might exhibit a cliff.

>>>>

The preceding discussion raises the issue of explicitly duplicating structural information in the
independent and dependent variables. More specifically, both the dependent variable (i.e., the SALI
values) and the aggregated (topological) descriptors characterize the molecules' structure. One
might therefore ask whether models built using such data perform over-optimistically. Given the
nature of the descriptors used in the independent and dependent variables (the former being based on
multiple atom and bond features and the latter derived from purely topological paths), we believe
that such a phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram of the pairwise
Pearson correlations between the SALI values and each of the descriptor values for the
Cavalli\cite{Cavalli:2002aa} dataset, using the different aggregation functions described above. It
is evident, that the highest $R^2$, between any of the descriptors and the SALI values is less than
0.15. Similar behavior was observed for all the datasets used in this study. These observations
suggest that problems arising from the inclusion of structural information simultaneously in the
dependent and independent variables is minimal.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values using models built
with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs $f_{\textrm{absdiff}}$, etc.)
indicated that the differences between the sets of predictions was statistically insignificant (all
$p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
\ref{tab:rfmodel-results} reports the model statistics.  While the differences in model performance
do not vary significantly with the aggregation function used, we see that for the Costanzo and Kalla
datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and $R^2$ than when
using $f_{\textrm{mean}}$.

Given this observation, we arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function
for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}
summarize the performance of the models built using the reported activity data and the pairwise SALI
values for the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI
models exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to
perform better at the higher end of the SALI values. However in all three cases, the most
significant cliffs exhibit significant variance. In comparison, the models developed on the reported
activity values are relatively poor. This can be partially ascribed to the small sizes of the
dataset. It should be note that this methodology is not aimed at modeling the original, usually
small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
not lead to very robust or reliable machine learning models. Instead, the methodology focuses on the
\emph{pairwise} dataset -- which are usually large enough for a statistically significant model.

Since the model is built on pairs of observations, performance measures characterizing the model are
based on the accuracy of pairwise SALI values. The model can also be tested in a slightly different
approach by checking whether it can correctly identify cliffs between a new molecule and the member
of the training set. To test this, we consider the original set of $n$ individual molecules and hold out
$m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We then
evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. The
model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
remaining molecules.
